---
title: "The Impact of War on the Economies of Russia & Ukraine"
subtitle: "Proposal"
author: 
  - name: "Stock Sleuths"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Our team is conducting a comprehensive analysis with the goal of identifying and analyzing trends in the financial markets of both Russia and Ukraine, as well as their allies. By examining these trends, we aim to gain valuable insights into how the ongoing conflict has influenced the economies of the two countries and their allies."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

## Questions

The two questions you want to answer.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
