---
title: "The Influence of Russia/Ukraine Conflict on US Economy"
subtitle: "Proposal"
author: 
  - name: "Stock Sleuths"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "We aim to analyze the impact of the Russia/Ukraine conflict on the U.S. economy by focusing on oil prices and stock market trends from prior to the conflict, the beginning of the conflict (February 2022), and up to present day. By examining these trends, we aim to gain valuable insights into how the ongoing conflict has influenced the US economy, as an ally of Ukraine."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Datasets

```{r}
#| label: load-dataset
#| message: false
```

We aim to analyze the impact of the Russia/Ukraine war on the U.S. economy by focusing on oil prices and stock market trends from prior to the conflict, the beginning of the conflict (February 2022), and up to present day. To do this, we will use three primary datasets: one from the U.S. Energy Information Administration (EIA) containing data on U.S. crude oil prices, a Kaggle dataset tracking S&P 500 stock market data, and a dataset from the Federal Reserve Economic Data (FRED) on U.S. consumer sentiment. 

These datasets will provide key insights into how these critical economic factors have shifted since the onset of the conflict, enabling a detailed exploration of potential correlations and trends. While the data is quantitative, we must consider if there might be underlying biases in the datasets that could skew results. For example, the consumer sentiment data could reflect certain demographic biases depending on how the data is collected, especially if it under represents certain groups of people.

**Dataset 1:** US Stock Market Data from Kaggle, sp500_stocks.csv (200 MB)  (<https://www.kaggle.com/datasets/andrewmvd/sp-500-stocks>) 

This dataset is updated daily and provides historical data, from as far back as 2010 up to current day, on the S&P 500, or Standard and Poor's 500, which is one of the world's most renowned financial benchmarks. This stock market index measures the performance of 500 major companies listed on U.S. stock exchanges, with the data originally being collected from FRED and Yahoo finance. Although it tracks 500 companies, the inclusion of multiple stock classes from some firms, such as Alphabet's Class A (GOOGL) and Class C (GOOG), means the index actually comprises 505 stocks. Since this data is aggregated and anonymized, there are no privacy concerns for individual investors.

**Dataset 2:** EIA data pswrgvwall.xls (7 MB) <https://www.eia.gov/petroleum/gasdiesel/>

This dataset is from the U.S. Energy Information Administration and it provides historical data for crude oil prices in the US (spanning from the 1990s to present day), which has correlation with geopolitical events that affect supply and demand. It comes from a reputable government source, which means the data is considered reliable and transparent. However, it's always essential to acknowledge any potential limitations or assumptions made in the data collection process by this agency, as such limitations could influence analysis and its conclusions.

**Dataset 3:** Consumer Sentiment & Inflation data from FRED umcsent.csv (501 bytes) and MICH.csv (8 KB) -merged together- <https://fred.stlouisfed.org/series/UMCSENT> <https://fred.stlouisfed.org/series/MICH>

These datasets are from the Federal Reserve of Economic Data and provide data on economic indicators such as consumer sentiment and inflation expectations, to show the influence of economic changes on the economy and how society reacts. It is updated monthly and for the purposes of our project, we pulled data from February 2022 to present day. The data also comes from a trusted government source, which generally makes it reliable and transparent. However, it's important to recognize any potential limitations or assumptions in the agency's data collection process, as these factors could impact the analysis and its conclusions.

We chose these datasets and this topic because some of our team members have a strong interest in how data science can be used in the financial industry. We believe it will be insightful to analyze the impacts of the Russia/Ukraine conflict on the US economy, specifically on stocks and oil prices. Our team agreed exploring this relationship would be a valuable and creative way to strengthen our data visualization skills.

## Questions

**Question 1: How have US stock and oil prices changed since the beginning of the Russia/Ukraine Conflict?**

**Question 2: How has consumer sentiment, stock prices, oil prices, and inflation expectations changed over time during the Russia/Ukraine conflict?**

## Analysis plan

**Question 1 Approach:** We will first analyze how U.S. stock prices and oil prices have evolved since the beginning of the Russia/Ukraine conflict. We will visualize these trends using a **time series plot** and explore the relationship between stock and oil prices through a **scatter plot**. Before proceeding with the analysis, we will clean and preprocess the data to address any missing or inconsistent values, ensuring the datasets are ready for analysis. We will also conduct exploratory data analysis (EDA) to identify initial patterns and trends in the data. 

**Variables**:

-   **U.S. stock prices** (quantitative)

-   **Oil prices** (quantitative)

-   **Time since the start of the conflict** (categorical or quantitative depending on how time is measured, e.g., by year/month or as a binary pre-/post-conflict variable)

This question is important because the Russia/Ukraine war has had significant global economic ramifications, particularly on energy markets and financial stability. Understanding the specific changes in U.S. oil and stock prices during this period can reveal the direct economic impact of geopolitical conflicts on the U.S. economy.

**Question 2 Approach:** We will investigate how consumer sentiment has fluctuated alongside changes in stock and oil prices. This will involve using a **box plot** to capture the distribution of consumer sentiment and a **line plot** to reveal trends over time. By combining these visualizations with statistical analysis, we aim to uncover meaningful patterns and connections between the conflict and U.S. economic indicators.

**Variables**:

-   **Consumer sentiment** (quantitative)

-   **U.S. stock prices** (quantitative)

-   **Oil prices** (quantitative)

-   **Inflation Expectations** (quantitative) – from the FRED dataset

This question is crucial because consumer sentiment is a key indicator of economic health, reflecting public confidence in the economy and influencing spending and investment decisions. By examining how sentiment has shifted in relation to stock and oil price fluctuations, we can assess the broader psychological and behavioral impact of economic changes triggered by the conflict. By including inflation expectations, we aim to understand not only how people feel about the economy (consumer sentiment) but also how they perceive future inflationary pressures, giving a more rounded view of economic sentiment during the war. Understanding this relationship helps in anticipating consumer responses to market volatility, which is valuable for businesses, policymakers, and investors looking to gauge the potential economic trajectory during crises.

\
